\textbf{Stability, Assurance, and the Concept of Legal Guidance}

\emph{\textbf{Abstract}}

\begin{quote}
Legal theorists standardly hold that stability is one of eight necessary
conditions for legal guidance. We lack an adequate explanation, however,
of why, exactly, stability is necessary in order that law possess the
capacity to guide behavior. Standard explanations, which rely on a claim
about reasonable expectations, fail to connect the concepts of stability
and legal guidance. In this paper, I argue that, according to the
leading conception of legal guidance, stability is, in fact, not
necessary in order for law to possess the capacity to guide. In
response, I provide an alternative conception of legal guidance, and
argue that stability \emph{is} a necessary attribute if we assume this
conception of guidance. More generally, the paper sets out an abstract
framework in which claims that this or that property is necessary for
legal guidance can be assessed.
\end{quote}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Introduction}
\end{enumerate}

Legal positivism is, in the first instance, a thesis about the
conceptual foundations of legal validity.\footnote{\emph{See},
  \emph{e.g.}, Mark Greenberg, \emph{How Facts Make Law}, \textsc{Legal
  Theory} 10(3)(2004)157:198; John Gardner, \emph{Legal Positivism: 5 ½
  Myths}, A\textsc{m. J. Juris}., 46(1)(2001): 199--227; Christopher
  Kutz, \emph{The Judicial Community}, \textsc{Phil. Issues}
  11(1)(2001): 442-469; David Enoch, \emph{Reason-Giving and the Law},
  in Leslie Green and Brian Leiter, eds., \textsc{Oxford Studies In
  Philosophy of Law}, Volume 1 (Oxford: Oxford University Press, 2011):
  1-15.} It seeks to parse the \emph{social} mechanisms that underlie
the \emph{legal} decisions that, together, determine the substance and
govern the processes of a given legal system.\footnote{\emph{See} Gerald
  J. Postema. \emph{Coordination and Convention at the Foundations of
  Law}, JOURNAL OF LEGAL STUDIES, 11 (1982): 165--203, at 167 (``Thus
  positivists in the tradition stemming from Bentham locate the
  {[}criteria of validity{]} in matters of social fact, thereby
  rejecting the view that the validity of a law is a function of its
  truth or moral soundness.''); \emph{see also} Andrei Marmor, \emph{The
  Rule of Law and its Limits}, \textsc{Law and Philosophy}, 23(1)(2004):
  5.} Natural law approaches, by contrast, seek to parse not only the
social but also the \emph{moral} mechanisms that underlie legal
determinations of validity.\footnote{\emph{See} Kutz, \emph{supra} note
  1, at 444-45. John Finnis, \emph{On the Incoherence of Legal
  Positivism}, 75 \textsc{Notre Dame L. Rev.} (2000): 1597-98.} For both
positivists and natural lawyers, proposition \emph{P} accurately states
the law governing some topic in a given jurisdiction, if and only if
\emph{P} meets the applicable criteria of legal validity, \emph{C}, in
that jurisdiction.\footnote{Kutz, \emph{supra} note 1, at 443.} The
fundamental difference between the two approaches is that positivists
maintain that \emph{C} obtains in only virtue of social mechanisms,
while natural lawyers remain open to the possibility that \emph{C}
obtains in virtue of moral mechanisms.

For positivists, the social mechanisms at the foundations of the law are
explicated through \emph{descriptive} facts: claims about the behaviors
or mental states of judges, legislators, executives, or
voters.\footnote{Greenberg, \emph{supra} note 1, at 157. For the classic
  statement, \emph{see} H.L.A. Hart, \emph{Positivism and the Separation
  of Law and Morals} 71 \textsc{Harv. L. Rev.} 71(4)(1958): 593-629. As
  Ronald Dworkin put it, for positivists, legal validity is a matter
  ``not {[}of the{]} content {[}of norms{]} but with their
  \emph{pedigree} or the manner in which they were adopted or
  developed.'' Ronald Dworkin, \textsc{Taking Rights Seriously}
  (Cambridge: Harvard University Press, 1978): 17.} So, for instance,
positivists may point to facts about the behavior of courts to explain
the content of \emph{C}, but not to the merits of \emph{C} itself.
Positivists, in other words, reject grounding their explanations of the
criteria of legal validity in \emph{value} facts: normative or
evaluative claims.\footnote{Greenberg, \emph{supra} note 1, at 157.
  There is, of course, the matter of hard and soft positivism. Soft
  positivists allow that value facts can possess derivative significance
  in determining the content of the law. Hard positivists argue that
  value facts can never do even that. The debate over whether value
  facts can affect the content of laws, not whether they can affect the
  criteria of legal validity.} These are taken to bear only an arbitrary
relationship with legal norms.\footnote{\emph{Id}. at 157-59.
  Positivism's insistence on the primacy of descriptive facts gives rise
  to a thorny question: How can facts about what we happen to have been
  doing in the past give rise to genuine reasons about what we should be
  doing in the future? Or, put differently, how can it be that what we
  happen to do around here can provide reasons to keep doing it?
  \emph{See}, \emph{e.g.}, Kutz, \emph{supra} note 1, at 446.} Natural
lawyers, by contrast, might well explain the content of \emph{C} through
reference to value facts. Perhaps, for instance, \emph{C} states the
criteria of legal validity because it is demanded by democratic values.
Explanations of these sorts are not open to positivists, for whom
descriptive facts possess explanatory primacy. In giving descriptive
facts explanatory primacy, positivists are committed to the claim that
there is no law, properly so called, absent that which is created
through social mechanisms.

One of the most important of these social mechanisms -- and the focus of
our attention in this paper -- is the process by which legal officials
posit a law, and individuals respond. Say that a legislature duly enacts
proposition \emph{P}. What follows? What is the characteristically
\emph{legal} way in which individuals are influenced by \emph{P}? The
way in which one answers that question determines her conception of
\emph{legal guidance}.

A conception of legal guidance explains what attributes law must
possess, such that it has the capacity to guide or direct individual
conduct. Although positivism is, in the first instance, a thesis about
legal validity, it is not only that. It is also a descriptive theory
that seeks to explain how law, ``as something that must be
\emph{posited} through some social act or activity, either by enactment,
decision, or practice,''\footnote{Scott Shapiro, \emph{Law, Morality,
  and the Guidance of Legal Conduct}, \textsc{Legal Theory} 6(2)(2000):
  127 (emphasis in original).} possesses the capacity to guide or direct
individual conduct.

This question is particularly pressing for legal positivists. If a
natural lawyer claims that \emph{value} facts possess explanatory
primacy in an account of legal validity, such that, at least in some
cases, when \emph{P} requires one to ϕ, one \emph{thereby} has a genuine
or unqualified reason to ϕ, then the task of explaining the connection
between law and practical reasoning is, arguably, less urgent. For
natural lawyers, the connection between the content of \emph{P} and
individual beliefs and action is more straightforward. In short, if
\emph{P} is underwritten by some moral logic, it \emph{prima facie}
possesses the capacity to guide conduct.\footnote{There are a number of
  ways in which we might draw a link between natural law and the
  capacity to guide behavior. For instance, one might maintain that the
  capacity to guide conduct is a necessary feature of \emph{moral}
  norms. If all moral norms possess the capacity to guide, then any
  legal norm underwritten by a moral norm will possess the capacity by
  extension.} Positivists, however, have nothing of the sort upon which
to rely. They must, instead, explain how it is that laws, which are not
necessarily underwritten by any moral logic, possess the capacity to
guide individual behavior.

The centrality of legal guidance has not eluded positivists. Scholars
working in the field have developed a comprehensive conception of legal
guidance and set out the social and legal conditions under which it
obtains. The conventional wisdom conceptualizes legal guidance as that
state of affairs in which an individual ``is able to learn of his
obligations or rights from {[}the relevant legal norm{]} without
engaging in deliberation.''\footnote{\emph{See} Shapiro, \emph{supra}
  note 8, at 153.} This conception of legal guidance treats guidance as
a purely \emph{epistemic} phenomenon. Law guides when it possesses the
capacity to \emph{convey information} about the content of the law to
individuals. In order to possess this capacity, positivists hold,
``there are certain conditions that the law has to meet.''\footnote{Marmor,
  \emph{supra} note 2, at 5.} A wide consensus holds that eight
attributes ``state necessary conditions'' \footnote{Scott Shapiro,
  \textsc{Legality} (Cambridge: Harvard University Press): 395. Earlier,
  Shapiro observes that ``regimes that flout these principles are simply
  not engaged in the basic activity of law.'' \emph{Id}. at 394.} for
the existence of legal guidance and thus legality. These attributes are,
by now, familiar. Laws must be general; they must be promulgated; laws
must not succumb to retroactivity nor contradiction; they must be clear,
and possible to follow; there must be congruence between the rules as
written and as applied; and, finally, laws must be stable. That legal
norms must possess each of these attributes in order to guide -- in the
sense of conveying to individuals their legal rights and obligations --
is conventional wisdom among legal positivists.\footnote{\emph{See
  generally} H.L.A. Hart, \textsc{The Concept of Law} (2)(Oxford: Oxford
  University Press, 1997); Joseph Raz, \emph{The Rule of Law and its
  Virtue}, in \textsc{The Authority of Law} (1979); Jeremy Waldron,
  \emph{The Concept and the Rule of Law}, \textsc{Harv. J. L. \& Pub.
  Pol'y.} 30(2006):15-30; John Gardner, \emph{Hart on Legality, Justice,
  and Morality}, \textsc{Jurisprudence} 1(2)(2010): 253-265.}

This paper challenges the conventional wisdom about epistemic guidance
and the general picture painted by positivists explaining how
individuals are informed about and motivated by legal norms. It does so
through the lens of an alleged attribute that has proven difficult to
explain\footnote{\emph{See}, \emph{e.g.}, Marmor, \emph{supra} note 2.}
yet nonetheless is held essential for epistemic guidance: the stability
requirement, which holds that laws should possess a certain ``constancy
through time'' and ``not be changed too frequently.''\footnote{Lon
  Fuller, \textsc{The Morality of Law: Revised Edition} (New Haven: Yale
  University Press, 1969): 79.} This paper develops a means for testing
whether any given attribute of guidance is, in fact, necessary for law
to possess the capacity to guide, and then applies that test to the case
of stability. The paper argues that, under the standard conception of
epistemic guidance, stability is not, in fact, required to guide
conduct.\footnote{Thus, while the paper is avowedly abstract in its
  primary concerns, it possesses implications for public policy. The
  conventional wisdom argues that legal norms must meet a certain set of
  parameters if they are to possess the capacity to guide conduct. This
  is a straightforwardly descriptive claim. These parameters place
  constraints on the form legal norms can take. If they are true, they
  thus limit the substantive policies the state can pursue, assuming the
  state wants to use law to guide behavior. If we think of law as a
  technology or tool for governance, the eight attributes of guidance
  essentially state the limitations of the tool. Because the criteria of
  legality place substantive constraints on policy, they potentially
  exclude democratically selected policies.} Yet, legal theorists have
long linked stability with guidance. In an effort to rationalize this
long tradition of linking stability with guidance, this paper presents
an alternative account of legal guidance. Rather than conceptualize
guidance as purely epistemic, the paper presents a \emph{motivational}
account of guidance. Whereas an account of epistemic guidance explains
merely how law conveys its requirements, an account of motivational
guidance explains how law could be \emph{the reason why} individuals
comply. Such an account, the paper argues, is valuable not only because
it best explains legal theorists' long-standing tendency to link
stability with guidance, but also because it offers an account of the
way in which the state and its citizens cooperate to overcome the harm
to reliance interests possibly brought about by unstable law. The
account of motivational guidance presented here models the state and its
citizens as playing an assurance game, and argues that legal change,
even change that is responsive to widely-held problems, will tend to
only occur when both sets of parties are sufficiently assured that the
other will cooperate. We begin our argument, however, with a much more
standard idea: the way in which law, as a descriptive matter, is able to
shape individual deliberation.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Epistemic Legal Guidance: Mediating between Law and
  Deliberation}
\end{enumerate}

When the U.S. federal government enacts new criminal laws, or seeks to
change health care policy, or changes the rules governing securities
laws, it seeks to regulate the behavior of hundreds of millions of
individuals.\footnote{\emph{See} Shapiro, supra note 12, at 72
  (describing ``the regulation of mass populations''). \emph{See also}
  Jeremy Waldron, \emph{The Rule of Law in Contemporary Liberal Theory},
  2(1) \textsc{Ratio Juris} (1989): 79-96. The basic demographic
  contours of the U.S. population are available at United States Census
  Bureau, www.census.gov. It should be noted that U.S. law, in addition
  to governing those within the territorial United States, also applies
  to certain individuals outside its borders.} Regulation at this scale
poses a problem: how is the state to \emph{inform} this number of
individuals of their new duties, rights, or obligations? Even if each
individual governed by a new regulation could be located, it is clearly
beyond the capacity of the state to send officials door-to-door
providing updates on new regulations.\footnote{Although, on certain
  occasions, the state does organize events, like community workshops,
  to help inform individuals of their new duties, rights, and
  obligations.} As a substitute, the state crafts legal norms that
provide citizens with \emph{epistemic guidance}. Laws provide epistemic
guidance when a citizen ``is able to learn of his obligations or rights
from {[}the relevant legal norm{]} without engaging in
deliberation.''\footnote{\emph{See} Shapiro, \emph{supra} note 8, at
  153.} More colloquially, laws provide epistemic guidance to
individuals when an individual is able to read the law and thereby learn
how to act in compliance with the law. Such \emph{impersonal} guidance
is a substitute for the \emph{personal} guidance rendered infeasible by
the size of the modern state and its large-scale regulatory
schemes.\footnote{On occasion, law is used, of course, to regulate
  single individuals or small groups. Legislatures pass bills aimed at
  single individuals. \emph{See}, \emph{e.g.}, Jeffrey S. Hill and
  Kenneth C. Williams, \emph{The Decline of Private Bills: Resource
  Allocation, Credit Claiming, and the Decision to Delegate}, 37(4)
  \textsc{Am. J. Pol. Sci. (1993): 1008-1031.} Judicial decisions affect
  only the parties to the case, in the first instance. And
  administrative agencies adjudicate claims of single persons.}

Epistemic guidance offers a solution to the problem of regulating
large-scale populations. But not all laws provide epistemic guidance. A
wide consensus holds among legal positivists that epistemic guidance is
possible \emph{only} if the legal norms that a state promulgates meet
certain parameters. In order to guide epistemically, positivists hold,
``there are certain conditions that the law has to meet.''\footnote{Marmor,
  \emph{supra} note 2, at 5.} Let us call these conditions -- which
positivists argue ``state necessary conditions'' \footnote{Scott
  Shapiro, \emph{supra} note 12, at 395. Earlier, Shapiro observes that
  ``regimes that flout these principles are simply not engaged in the
  basic activity of law.'' \emph{Id}. at 394.} for the existence of
epistemic guidance -- the \emph{attributes of guidance}.

Conventional wisdom among positivist legal philosophers holds that, in
order to guide individual conduct, ``there are certain conditions that
the law has to meet.''\footnote{Marmor, \emph{supra} note 2, at 7.} A
wide consensus holds that there are eight attributes that ``state
necessary conditions for the existence of {[}law{]}.''\footnote{Shapiro,
  \emph{supra} note 12, at 395.} We need only briefly review the
substance of these conditions. Laws must be (1) \emph{general}, setting
forth rules of conduct, applicable to some defined segment of the
population, prohibiting or facilitating certain modes of behavior. These
rules must be published, or (2) \emph{promulgated}, so that they are
available to those whose conduct they govern on a (3)
\emph{prospective}, rather retrospective, basis. Laws must avoid
obscurity or unintelligibility; they must be (4) \emph{clear}. A legal
code must be (5) \emph{non-contradictory}, in that its provisions do not
conflict. Closely related is the demand that law be (6) \emph{possible}
to follow. Official enforcement of these laws should be (7)
\emph{congruent} with the laws as written. And, finally, laws should be
(8) \emph{stable}, which requires that norms not change too frequently.
As a knife must be sharp if it is to cut well, so too must laws meet
these criteria if they are to provide epistemic guidance.\footnote{Raz
  helpfully compares law to a knife, in the sense that both can possess
  certain properties that render them, as tools deployed by individuals,
  more or less useful. Our question can usefully be framed: If sharp
  knives cut well, what sort of laws govern conduct well? \emph{See}
  Raz, \emph{supra} note 13, at 225.}

We, of course, are concerned here with the attribute of stability in
particular. Ordinarily one might pause here and explicate the idea of
stability in some detail. The difficulty is that positivists have
remained rather coy about just what stability amounts to. Andrei Marmor
captures contemporary sentiment about stability: ``This requirement of
the rule of law is basically a rough standard,'' argues Marmor, because
``it would be absurd to assume that we can have a precise notion of the
ideal pace of change.''\footnote{Marmor, \emph{supra} note 2\textsc{,}
  at 34. It is instructive to observe that Marmor suggests that, in
  discussing stability, we are searching for an ``ideal'' pace of
  change. This interpretation of the attributes of guidance seems overly
  moralized. The items on the list explain how laws should be crafted if
  they are to guide individuals; they do not instruct officials how to
  make ideal policy.} Our inability to formalize the stability
constraint seems to stem from the fact that, unlike other attributes of
guidance, it is difficult, if not outright impossible, to define
stability in the abstract. To characterize a law as stable seems, in
some fundamental way, to involve an assessment of the conditions in a
society, and the expectations of the governed, in a way that the other
items on the list do not. In the second half of this paper I will offer
a conception of motivational guidance that cashes out stability in just
these terms -- but prior to reaching that point, we will need to
investigate the conventional approach to stability.

The conventional wisdom holds that, if a legal norm meets the criteria
stated by the attributes of guidance, then it provides epistemic
guidance and, thus, is capable of \emph{making a difference} to
individuals' beliefs and conduct -- ``a difference, that is, in the
structure or content of deliberation and action.''\footnote{Jules
  Coleman, \emph{Incorporationism, Conventionality, and the Practical
  Difference Thesis}, 4(4) \textsc{Legal Theory (1998): 383.} We can
  test whether or not an individual was guided by a rule by asking about
  the counterfactual case: A legal rule, \emph{P}, guides a person to do
  some act, \emph{A}, only if that person ``might not have done''
  \emph{A} ``had he not appealed to'' \emph{P} qua legal rule.
  \emph{See} Shapiro, \emph{supra} note 8, at 132.} Epistemic legal
guidance, then, and the attributes of guidance that make it possible,
provide a crucial explanatory link for positivists. It explains how
laws, which begin as strings of characters posited by officials, govern
society.

Crucial to this explanatory chain is the idea that individuals, once
they learn of their rights and obligations from law, take the content of
law as reasons for action.\footnote{Enoch, \emph{supra} note 1; Gerald
  J. Postema, \emph{Implicit law}, LAW AND PHILOSOPHY, 13(3)(1994):
  361--387; Stephen Perry, \emph{Hart's Methodological Positivism}, in
  HART'S POSTSCRIPT: ESSAYS ON THE POSTSCRIPT TO THE CONCEPT OF LAW
  (Oxford: Oxford University Press, 2001).} Law transmits its directives
by providing individuals with things called reasons.\footnote{Pamela
  Hieronymi, \emph{Reasons for Action}, \textsc{Proceedings of the
  Aristotelian Society}, 111(2011): 407--427.} Individuals, in turn, are
equipped to process and make sense of these reasons. Although
philosophers dispute the exact set of processes that individuals use to
process reasons -- is reasoning machine-like or driven in part by
emotions? -- it is generally agreed that the processing of reasons
involves some, most likely imperfect, process of \emph{weighing}.
Discrete considerations can be measured against each other, and added
and subtracted, in order to help individuals arrive at this or that
belief. Hence we often speak of a \emph{balance} of reasons. Legal norms
provide epistemic guidance when they possess the capacity to add a
reason to one side or the other of the balance.

At this point one might well wonder, when we speak of ``legal norms,''
if we are referring to all legal norms, or only a subclass of them. One
could, following Joseph Raz, argue that only the legal norms governing
the \emph{processes} that produce legal rules must be stable, not the
legal rules themselves.\footnote{Raz, \emph{supra} note 13, at 214-216.}
Guidance, Raz argues, requires merely that ``the making of particular
laws should be guided by open and relatively stable general
rules.''\footnote{\emph{Id}. at 215.} What distinguishes a general rule?
Raz specifies that ``two kinds of general rules create the framework for
the enactment of particular laws,'' namely, ``those which confer the
necessary powers for making valid orders'' and ``those which impose
duties instructing the power-holders how to exercise their
powers.''\footnote{\emph{Id}.} Under Raz's framework, general rules
govern the creation of the primary rules that regulate individual
conduct. And it is general rules that must be stable in order that law
guide conduct.

Notice, however, that it is also the case that any given process rule
can be re-described as a discrete rule that applies to some legal
official.\footnote{Ultimately, some official enforces the rules that
  govern process.} All process rules are discrete rules, in this sense.
Thus, any argument we can run to show that stability is not a necessary
condition for epistemic guidance works for both discrete norms and
process rules. Insofar as stability is concerned, Raz's distinction
doesn't make a difference.\footnote{Raz's reason for drawing this
  distinction is well taken. Raz is concerned that if we do not draw the
  distinction, most of administrative law will fail to be counted as
  law, since it is notoriously volatile. Raz singles out administrative
  regulations as the site at which stability of general legal processes
  are of particular import for the stability requirement. We will show
  later in the argument that one can sustain administrative law as law
  without resorting to the discrete rules versus process rules
  distinction.}

Moreover, instability of this sort -- instability so pervasive that it
affects the conventions sustaining the creation of laws -- is the type
of instability that undermines an entire legal system. Legal systems
tend not to survive such dramatic shake-ups. And legal systems, as such,
are standardly defined as stable entities.\footnote{Shapiro,
  \emph{supra} note 12, at 65.} If Raz's distinction \emph{does} make a
difference, it proves too much. Thus, absent a compelling reason to
distinguish types of legal norms, we shall use the term to refer
indiscriminately to both general and primary rules.

It \emph{is} important, however, to distinguish between legal and
non-legal norms. Unsurprisingly, legal norms do not monopolize
individuals' deliberations. \emph{Non-legal} reasons compete with legal
reasons for individuals' allegiance. In contemporary states, individuals
receive guidance not only from legal norms, but also from customary
norms, ethical norms, religious norms, and social norms. Each of these
types of norms might \emph{also} purport to regulate individual conduct.

The state assists individuals in locating the \emph{legal} norm by
adorning it with ``an official marking that designates certain standards
as those to which one must conform.''\footnote{Jules L. Coleman,
  \textsc{The Practice of Principle: In Defense of a Pragmatist}

  \textsc{Approach to Legal Theory} (Oxford: Oxford University Press,
  2003): 137.} Common marks include ``inscription in some authoritative
text'' or ``declaration by some official.''\footnote{Scott Shapiro,
  \emph{What is the Rule of Recognition (and Does it Exist?)},
  \textsc{Yale Law School Public Law and Legal Theory Research Paper
  Series} (unpublished manuscript): 2, available at
  \emph{http://papers.ssrn.com/sol3/papers.cfm?abstract\_id=1304645}.}
Official marks allow citizens to identify that a norm is, in fact, a
legal norm without having to engage in deep deliberation and speculation
as to the pedigree of the norm. For positivists, a legal norm is valid
just in virtue of certain social facts, usually certain political
procedures, such as passage of bills, and the like. Adorning legal norms
with official marks is designed to distinguish legal norms from other
sorts of norms, and to ensure that individuals need not engage in the
deep historical and conceptual tasks that would be involved in an
investigation of whether a given norm is, in fact, a legal
norm.\footnote{\emph{See} Hart, \emph{supra} note 13, at 95.} Official
marks attempt to solve the informational problem that arises when
competing norms purporting to regulate proliferate. An official legal
mark conveys to the individual the idea, as far as the law is concerned,
\emph{this} norm is dispositive.\footnote{Nonetheless, an individual
  might wonder why comply with the legal norm rather than some other
  norm. Official markers do not provide an all-things-considered reason
  to comply with the legal standards, but, in designating certain
  standards as those with which law requires conformity, official
  markers convey that legal reasons do not take themselves as subject to
  the balance of total reasons. Official markers demonstrate that legal
  reasons, simply in virtue of their status, not only are not subject to
  the balance of all reasons, but take themselves to have changed the
  balance of reasons in a Razian manner or do not take themselves to
  have changed the balance of reasons but do simply disregard the other
  reasons.}

Epistemic guidance, therefore, involves \emph{presenting} individuals
with the legal norm alongside whatever non-legal norms to which they
adhere. In other words, epistemic guidance models a \emph{thin}
integration between the reasons created by legal norms and an
individual's deliberative processes. Individuals pursue their
antecedently chosen ends, and, in the pursuit of those ends, encounter
legal norms. Background beliefs and desires determine whether, and to
what extent, individuals comply with legal norms. Legal norms provide
something of a map of possible routes for individuals, assuming that the
individual knows her destination. Which route any given agent takes is a
function of antecedently given desires. And the individual may choose to
simply not consult the legal map at all, depending on these antecedently
given desires. Epistemic guidance does not require that individuals
choose their ends because of legal norms -- only that individuals can
make themselves aware of what the law requires.\footnote{Here is another
  way of putting the point. Assume an individual with a set of beliefs,
  B, about what the law requires. Artificially separate the individual's
  desires into some set of desires toward compliance with the law, D,
  and all of her other desires, O. Individuals' deliberations concerning
  the pursuit of O are constrained by the interaction of B and D. So,
  e.g., if the individual wants to get rich, but also wants to comply
  with the law, then things like robbing a bank will be ruled out. Now,
  in the ordinary course of things robbing a bank will also be ruled out
  by moral or social desires. In such a case, that the individual
  doesn't rob a bank is overdetermined. But let's assume that, for this
  individual, moral or social pressure doesn't rule out robbing a bank;
  the desire to comply with law is the only constraint. Say B changes,
  to B-prime. Now, O is constrained by B-prime. O does not change; only
  the content of the constraint is changed.} This would seem to be the
case even on a Razian account premised on peremptory reasons.\footnote{\emph{See}
  Joseph Raz, \emph{Authority, Law and Morality}, 68(3) \textsc{The
  Monist} (1985): 295-324. I am very grateful to an anonymous reviewer
  for raising this objection.} For even if it is the case that law
presents itself as a practical authority, and is ``meant to replace the
reasons on which it depends,'' it still may be the case that the citizen
determines that the reputed practical authority is no authority at
all.\footnote{\emph{Id}. at 297.} By \emph{the rules of the legal game},
legal norms purport to be practical authorities; but epistemic guidance
does not require that citizens themselves conclude, as an all-in
judgment, that legal rules are always authoritative.\footnote{As Raz
  writes, ``No blind obedience to authority is here implied. Acceptance
  of authority has to be justified \ldots{}'' \emph{Id}. at 299.
  Epistemic guidance, for Raz, requires that law ``must be capable of
  guiding the behavior of its subjects. It must be such that they can
  find out what it is and act on it.'' Raz, \emph{supra} note 13, at
  214.} If it did, it would imply that epistemic guidance is only
satisfied when all citizens comply all the time. All that epistemic
guidance requires is that legal norms be written in such a way as to
convey their normative demands to citizens.

The conventional wisdom holds that epistemic guidance is necessary in
order that individuals be made aware about what the law requires of
them. In a world in which personal guidance is infeasible, individuals
learn of their rights and obligations impersonally, through epistemic
guidance. Because individuals' beliefs and actions are determined by
weighing reasons, legal norms affect beliefs and action by creating new
reasons for individuals. Thus, on this descriptive model, legal norms
operate via ``self-directed action.''\footnote{Postema, \emph{supra}
  note 28, at 369.} In other words, individuals must apply general legal
directives to their own particular situations, thus overcoming the
problem of mass regulation. Law thus guides by seeking to affect
individuals with reasons.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Does Epistemic Guidance Require that Law be Stable?}
\end{enumerate}

In order to guide by reason -- in order to provide the epistemic
guidance crucial for governing mass populations -- must legal norms, in
fact, meet the criteria set forth by the attributes of guidance? Given
this picture of epistemic guidance, which mediates between laws on the
books and individuals' deliberations and actions, is it, in fact, true
that the attributes of guidance state necessary conditions for epistemic
guidance? More formally, we may ask of any of the eight attributes:

\begin{quote}
What state of the world does epistemic guidance require such that some
attribute, \emph{X}, is necessary, assuming the operation of some other
set of attributes, \emph{Y}, where \emph{X} ∉ \emph{Y}?
\end{quote}

We answer this question by posing a counterfactual. An attribute is
necessary for a conception of legal guidance when the absence of that
attribute renders law incapable in principle of guiding individuals. The
idea here is that there are certain states of the world that must obtain
in order that legal guidance be possible, and the items constituting the
attributes of guidance either uniquely cause those states or are
necessary for those states to exist. Posing a counterfactual in which
legal norms do not possess the attribute in question allows us to test
whether guidance is possible in its absence. Applying this framework to
the case of stability, we can generate two cases: one in which a legal
norm is stable and a counterfactual case in which a legal norm is
unstable. We then compare how these two cases affect epistemic guidance.

First, imagine a successful case of epistemic guidance, in which the
applicable law, \emph{P}, is stable. Say that Alice wants to invest some
money in a stock that will grow in value over time. Alice is indifferent
between stocks, so long as the one she picks will grow in value. One
investment option is a company's stock that Alice read about in the
newspaper. Another option is a company's stock that her friend, who
possesses insider information, told her would grow in value because it
is going to be acquired by another company shortly. Prior to consulting
the applicable law, Alice is indifferent between the two options, since
all she wants to do is invest in a money-making stock. However, because
Alice is motivated to comply with law, she seeks epistemic guidance.
Alice first seeks indirect guidance -- from her associates, secondary
sources, and so forth -- and receives a uniform set of reports as to the
content of the law. They tell her that trading stocks on insider
information is illegal, and could land her in jail. Just to be sure,
Alice reads the applicable law and applicable judicial opinions, and
finds that her indirect guidance is supported by the direct guidance.
One alternative, investing in the stock she read about in the newspaper,
is within the bounds of compliance, and the other alternative, trading
on insider information, is prohibited. Alice chooses the legal
alternative, and avoids jail.

Our successful case of epistemic guidance provides a baseline against
which we can pose a counterfactual. Can we alter the case such that
instability interferes with the transmission of legal content, thereby
impairing epistemic guidance? It is difficult to imagine such a case.
Even if the contents of \emph{P} changed at random intervals, Alice
could, in principle, locate the content of the law and thus receive
epistemic guidance. This conclusion follows from the definition of
epistemic guidance. No matter how volatile a discrete legal rule, if it
is duly enacted, then it bears law's official mark. And the fact that a
norm bears law's mark implies that the informational problem about which
norm is authoritative has been solved, and epistemic guidance is, in
principle, possible. Duly-enacted laws are published, of course, thanks
to the promulgation requirement. Laws altered in any random interval are
still subject to promulgation, and the fact of promulgation ensures that
individuals can in principle locate the content of the law, which is all
that epistemic guidance requires. Epistemic guidance simply does not
require any state of the world for which stability is required. Contra
the conventional wisdom, then, it seems that no level of instability
interferes with the capacity of legal norms to provide epistemic
guidance.

If our account thus far is correct, the conventional wisdom, which holds
that stability is a necessary attribute of epistemic guidance, is
incorrect. Where did those accounts go wrong? We need an error theory.
Here, we will explore why it is commonly believed that epistemic
guidance requires stability. The culprit is that instability raises the
costs of epistemic guidance. Instability makes it more difficult to
locate the law, thus making it more difficult to comply with the law. On
this line of thought, instability sets off a chain reaction that ends
with decreased compliance.

This line of thought dovetails with the most common way to justify
stability as an attribute of guidance. As earlier noted, the most common
way to do so is on grounds that, if law is volatile, individuals will
not rely on it. This claim, or some version of it, might well be true.
The problem with this argument is that epistemic guidance does not
require increasing individuals' reliance on law or otherwise providing
motivational guidance. It does not require legal officials to cultivate
reliance or that individuals be disposed to rely. If the attributes of
guidance \emph{are} conceptualized as attributes that tend to increase
guidance or compliance, then it would make sense to justify stability on
reliance grounds. The difficulty here is that if we conceptualize the
attributes of guidance in this way, then our list of eight attributes
is, and always was, woefully incomplete. If the standard is attributes
that tend to enhance compliance, then surely Pareto optimality, or
something along these lines, is missing from the list.

Nonetheless, one might still be worried about the possibility that high
volatility makes epistemic guidance more difficult, thus rendering
compliance less likely. This worry is not altogether misguided. Even if
stability is not required for epistemic guidance, it might possess a
contingent, yet predictable, relationship to compliance and thus law's
efficacy.\footnote{Ultimately, the level of compliance sought by legal
  officials is a function of the regulatory regime. Thus, higher or
  lower search costs might be tolerated. We cannot answer that in the
  abstract, so we cannot specify how much non-guidance is compatible
  with a regulatory program. We can only observe that there will be, in
  the normal case, some desired level of guidance and compliance.}
Sketching out such a contingent relationship does help to explain the
error, and explain why we might care about stability even if it is not a
property of legality. The key to sketching out this contingent
relationship between instability and reduced compliance is to treat
legal norms as costly information.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Legal Norms as Costly Information}
\end{enumerate}

Ideally, locating legal norms would be frictionless. The environment in
which legal norms operate, however, ensures that locating law's content
is rather complex. In contemporary states, individuals receive guidance
not only from legal norms, but also from customary norms, ethical norms,
religious norms, and social norms. Each of these types of norms purports
to regulate individual conduct. Laws can only claim primacy over other
types of norms if they make themselves known. Earlier, we observed that
official marks \emph{attempt} to solve the informational problem that
arises when competing norms purporting to regulate proliferate. Here, we
show that official marks are \emph{signals} from the state to citizens.
As signals, the information they provide is not always perfect. It is
sometimes noisy, especially when laws are unstable. This situation
raises the costs of obtaining legal information, and thus can lead to
reduced compliance.

One way in which individuals might determine if a norm is, in fact, a
legal norm is to engage in a historical inquiry into the pedigree of the
norm. This historical inquiry would involve questions like, \emph{Did
Congress pass this norm}?, \emph{Does the Constitution forbid norms of
this sort?}, \emph{Has any court modified the norm}?, and so forth. Such
investigations, however, are likely to result in confusion. Instead, we
need a proxy or a signal to stand in for all of the historical
investigation that goes into determining if a norm is, in fact, a legal
norm. The best proxy or signal of legality is that the norm ``bear the
mark'' of legal authority. When a legal norm bears the mark of legal
authority, it announces to individuals that it possesses the proper
pedigree.\footnote{One might be tempted to avoid taking the term
  ``mark'' literally, since, at base, to be lawful simply means to have
  the correct pedigree, to be a duly enacted law. But the official mark
  is not equivalent to (or even a logical extension of) the fact that a
  norm is duly enacted. The mark is a proxy designed to solve
  informational problems and decrease deliberation. Only duly enacted
  norms receive official marks, but we can imagine a case in which a
  duly enacted law does not, for whatever logistical shortcoming,
  receive a mark but is nonetheless law. A law that is not published
  would not receive a mark, for instance. If the official mark is to
  solve an informational problem about which norms individuals are
  supposed to comply, then we must take the idea of a mark literally.}
In theory, official marks guide individuals by signaling those norms,
and only those norms, that are sanctioned by the state. As a signal,
however, it is liable to convey incomplete or erroneous
information.\footnote{The legal philosophy literature has not addressed
  these shortcomings with official marks. It has treated the idea of
  official marks quite generously.} In such cases, epistemic guidance
suffers.

Signals can be erroneous because inscriptions in authoritative texts do
not \emph{stop} signaling if a law is superseded. A superseded statute
does not report that it has been superseded. A superseded statute does
not stop bearing the mark of law once it has been superseded. Ideally,
we would devise a technique to signal the content of the law that could
update itself. Official marks as they currently exist, however, are not
self-updating.\footnote{It is worth observing that the technology is
  readily available to make official marks self-updating, at least with
  regard to those official marks published electronically. A simple
  piece of code embedded in the metadata of the standing law might well
  be able to provide self-updating signals.}

One might be tempted to argue that a superseded statute bears an invalid
mark. But marks are not valid or invalid. Rather, they are accurate or
inaccurate, but there is no way of discerning whether they are accurate
or not from the mark itself. Thus, every change in the content of a law
results in the creation of some erroneous signals, unless inaccurate
signals are immediately destroyed. Frequently changing legal norms
results in a proliferation of inaccurate signals. If, for instance, an
official report of the law is printed once every year, and a given legal
norm is changed every year for 10 years, then there will exist nine
superseded reports, unless they are destroyed. These reports are
inaccurate signals of the content of the law.\footnote{Inaccurate, that
  is, assuming that the content of the change is not a reversion to some
  past norm.} Instability thus recreates the informational problem that
official marks were intended to solve. Superseded marks are, in essence,
competitor norms. They state norms that conflict with the state's
position on what counts as compliance with the law.

In addition to the possibility that an official mark provides erroneous
information, official marks might also, or alternatively, provide
\emph{incomplete} information. Although a legal norm contains much of
the information that is necessary to comply with law, it does not always
contain all of the information required for compliance. \emph{That}
information is located in interpretations of the norm, in judicial
opinions, and the like. In other words, an official mark is partial when
it points to a string of text, and refers to that text as law, when, in
fact, the text in question only conveys \emph{part} of the information
an individual needs to know in order to comply with the law. In the
United States, many official marks are partial, because some legal
content is contained in judicial opinions. For instance, much of the
content of the Eighth Amendment's prohibition on ``cruel and unusual
punishment,'' is contained not in the text of the law but, rather, in
the text of judicial opinions.

In theory, official marks, or signals, point to a string of text, and
tell the citizen: This string of text contains all of the information
you need in order to comply with the law. In practice, however, these
official marks are partial. They convey only some of the information
needed to comply with the law.\footnote{\emph{Cf}. \emph{Appalachian
  Power Co. v. E.P.A.}, 208 F.3d 1015, 1020 (D.C. Cir. 2000).}

If instability increases the erroneousness or incompleteness of official
signals, it increases what we can describe, following Jules Coleman, as
\emph{search costs}.\footnote{Jules Coleman, \textsc{Risks and Wrongs}
  (Oxford: Oxford University Press) \textsc{(2002): 125-131.}} If these
search costs are high enough, the efficacy of a particular legal norm is
reduced. Efficacy is a necessary condition of legal systems, although
not itself an attribute of legal norms. This line of thought does not
show that stability is an attribute of legality, but it may help us to
identify the point at which the costs of finding the law become such
that compliance is compromised to the point that legal norms are
rendered inefficacious.

Volatility can increase the costs of locating the law, thus ramping up
the costs of compliance. Volatility can artificially reduce compliance
by making it costlier than \emph{standing} law would imply that it is.
Assume some ``natural'' rate of compliance with a given law. The rate of
compliance is natural in the sense that it is a function of individuals'
background beliefs and desires. Volatility of the sort we have described
here reduces compliance below this natural rate, even though the content
of the standing norm is the same.

It's possible that instability creates costs sufficiently high that they
deter individuals from locating valid law. Locating valid law in the
face of instability can take time and effort, and can even result in
individuals mistakenly acting on what they believe to be legal reasons.
An example will illustrate how these costs could mount under non-ideal
conditions.

Imagine a variation on Alice's successful case of epistemic guidance.
Assume that Ben, acting in good faith, seeks epistemic guidance. The
content of the applicable legal norm is highly volatile, but Ben is
unaware of this fact. Ben seeks indirect guidance by consulting
secondary sources, which provide mixed reports. Some treatises report
that the law permits only option \emph{P}. Reports from those in a
related industry indicate that \emph{P} and \emph{Q} are permitted.
Sensing that something is amiss (Ben is not a legal official, but is
aware that the law is supposed to be clear and predictable), Ben seeks
direct guidance from what he believes to be an authoritative legal text.
But because the law is volatile, Ben encounters a superseded statute,
which, of course, does not report that it has been superseded. The
statute bears the mark of law, and indicates that both \emph{P} and
\emph{Q} are permissible. Because \emph{Q} is the preferable option all
things equal, Ben adds a consideration to his stack of reasons and acts
accordingly. In fact, Ben erroneously relied on a superseded statute.
The valid statute prohibits \emph{Q}. So Ben fails to comply.

The costs individuals are willing to incur in their search for epistemic
guidance presumably have limits. To be sure, the absence of epistemic
guidance does not imply a lack of compliance in every case. It might
well be the case that most individuals can comply without guidance. But
over the long haul compliance suffers when guidance decreases. And when
volatility increases, compliance decreases because individuals prefer to
save the costs to time and effort. If, at a certain point, the costs
become high enough and the content of the legal norm is non-intuitive
enough that individuals cannot figure out the compliant behavior absent
the norm, compliance could decrease to such a degree that the system can
no longer be described as efficacious.

A second way in which search costs can impair epistemic guidance is when
individuals erroneously rely on a superseded official mark. Here,
individuals desire to comply, and believe that they are complying, but
in fact fail to comply because instability decreases the strength of the
signal sent by officials to individuals. Unlike cases where individuals
decide to stop the search due to the costs of searching and bite the
bullet on compliance, here individuals believe that they have been
guided and are complying.

In our hypothetical, for instance, Ben \emph{took} himself to be relying
on a reason created by law. In fact, he erroneously relied on a
superseded official mark. In such cases, compliance suffers not because
of a decision that is too costly, but, rather, because volatility leads
to a situation in which individuals' \emph{operative} reasons -- the
reasons on which they, in fact, act -- are different from the reasons
that do, in fact, \emph{justify} the action from the perspective of law.
The mistaken non-compliance induced by volatility raises the question of
whether Ben was epistemically guided. We can see that there was no
reason for Ben to act in the way that he did, yet it seems as though Ben
acted for a reason.\footnote{\emph{See, e.g.,} Jonathan Dancy,
  \textsc{Practical Reality} (Oxford: Oxford University Press) 2002.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Three Types of Reasons}
\end{enumerate}

Was Ben epistemically guided when the reason on which he acted was, in
fact, not a reason? By way of analogy, consider a game of chess between
Beginner and Expert.\footnote{Here the account follows Pamela Hieronymi,
  \emph{supra} note 29.} If, say, Beginner is considering whether to
move a bishop vertically, he might consult the rules of chess and find
that bishops are only allowed to move diagonally. The rules of chess
tell Beginner that moving the bishop diagonally (or at least omitting to
move it non-diagonally) is an appropriate action. The rules
\emph{justify} the action.

It turns out that although moving the bishop diagonally was a valid
move, it was not a smart one, and Expert has placed Beginner in a very
precarious situation. Beginner sees that his one way out of Expert's
trap is to undertake what he believes to be a ``castle.'' Beginner
attempts to think through the rules of castling, wondering if he may do
so despite the fact that he had already moved his king. Beginner
mistakenly concludes that the rules of chess do, in fact, allow him to
castle despite already having moved his king. Here, Beginner acts for
what he took to be a reason, but, which, under the rules of chess, is
not actually a reason at all. We can call such reasons \emph{operative}
reasons.

Beginner erroneously believed he could castle after moving the king, due
to his lack of familiarity with the rules of chess and a general sense
of anxiety. These reasons \emph{explain} Beginner's choice. By
introducing a distinction between considerations grounded in a normative
system (justificatory reasons) and considerations that an agent took to
be grounded in a normative system (operative reasons), we can see why
Ben was not epistemically guided by the superseded statute.

Volatility may induce individuals to mistakenly believe that they are
being epistemically guided by the legal system when in fact they are
not. Ben, like Beginner, \emph{sought} to comply with the rules. The
non-compliance in both cases resulted from the complexity of the rules
relative to the individuals' capacities to comply. And while we can now
see how to avoid the awkward implication that Ben was in fact guided by
an invalid legal rule, the practical worry that volatility decreases
guidance remains.

We have our error theory, then. We have identified why one might be
worried by unstable legal norms, even if stability is not an attribute
of legality. The claim that stability is necessary appears reasonable if
one improperly assumes that stability is necessary in order to maintain
the efficacy of the legal system, rather than to facilitate legal
guidance. The core concern is that volatility reduces the strength of
the signal created by law's official mark. Without a strong signal,
ordinary folk will have difficulty locating the content of the law,
which leads to a decline in epistemic guidance, which reduces the
efficacy of the system. This is a serious concern, yet it does not
explain why stability is necessary for law to possess the capacity to
guide conduct. We are thus left with a dilemma.

This dilemma is this: We must reject \emph{either} the claim that
stability is an attribute of guidance \emph{or} the claim that epistemic
guidance is an accurate account of legal guidance. Stability is widely
thought to be an attribute of guidance, and epistemic guidance is the
most prominent conception of legal guidance. Perhaps the way out of this
dilemma is to modify our conception of legal guidance. Thus, we will
consider an alternative conception of legal guidance. On \emph{this}
conception, stability is, in fact, a necessary attribute of guidance.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Integrated Guidance}
\end{enumerate}

Integrated Guidance (``IG'') reconceptualizes law's guidance
function.\footnote{We find strong hints of IG-like positions in the
  literature, most notably in the work of later Scott Shapiro, but also
  in the work of Postema and Raz.} IG treats guidance as possessing both
epistemic \emph{and} motivational dimensions. In addition to providing
epistemic guidance, law motivates individuals by disposing them to
comply with it. IG models individuals as planning agents with the
capacity to incorporate legally sanctioned means in their plans. Law
guides when its norms, owing to their status as dispositive settlers of
normative controversies, exert pressure on individuals to use the norms'
content as the bases of plans. If the goal of individual deliberation is
to efficiently coordinate a large number of beliefs and desires, legal
norms should outperform any other type of norm, since no other type of
norm promises to be dispositive of controversy. Broadly speaking, this
process can occur via either of two routes: filtering and facilitating.
Legal norms that prohibit certain behavior filter out unsanctioned means
from individuals' plans. Legal norms that facilitate conduct offer
individuals opportunities to employ sanctioned means to achieve their
ends. The content of legal norms filters out some means, and creates
others. The result is a legally sanctioned set of means from which
individuals can choose. If individuals create higher order intentions
structured around this legally sanctioned set of means, then they will
be more disposed to comply with the law. Because plans require a certain
degree of stability, the means by which individuals accomplish those
plans must also possess a degree of stability. But we cannot simply
leave things there, since individuals will have different preferences
over reconsideration. We will attempt to sketch how much specificity is
required, given a population with certain preferences over
reconsideration.

IG is most naturally modeled around Michael Bratman's planning
conception of individual agency.\footnote{Michael E. Bratman,
  \textsc{Intentions, Plans, and Practical Reason} (Cambridge: Harvard
  University Press) 1987; Shapiro, \emph{supra} note 12.} Very roughly,
the idea is that individuals possess certain desires, or outcomes that
they hope to achieve. Most of the time, these desires are not
immediately satisfied; certain intermediate steps need to be taken to
render their realization more likely. The planning conception of agency
explains how we structure means and ends together to help us to achieve
our desires. Building on the planning conception of agency, IG explains
how, given the way in which individuals ordinarily structure their means
and ends, legal norms guide behavior. Individual action is primarily
governed by a set of hierarchical plans, to which one is non-trivially
committed, and by which one is guided in the making of more specific
decisions. The existence of such plans is essential, because we all face
severe resource limitations: limitations of time, energy, cognitive
capacity, and so forth. And, yet, despite these limitations we
nonetheless must coordinate a large set of diverse activities. Plans are
special sorts of individual intentions that help to overcome these
limitations.\footnote{Michael E. Bratman, \emph{Planning and the
  Stability of Intention}, \textsc{Minds and Machines}, 2(1992): 1--16.}
Individuals weave plans of various levels of generality and complexity
together in order to create an agenda for future action. Rather than
lurch from decision to decision, unguided by any background agenda,
individual action is designed to bring about the goals contained in the
plans, and do so efficiently. Inefficient means, or impossible means,
are ``filtered out.''\footnote{Michael E. Bratman, David J. Israel, and
  Martha E. Pollack, \emph{Plans and Resource-bounded Practical
  Reasoning}, \textsc{Computer Intelligence}, 4(1988): 350.} Individuals
are left with plans that exert strong, although not absolute, control
over any given decision that an individual makes. Plans can be
reconsidered, as we shall shortly see, but IG treats reconsideration as
the exception, not the rule. In case an individual does reconsider a
plan, she might reach for a new plan from her ``plan library,'' or a set
of beliefs about which actions bring about which effects under a given
set of circumstances.\footnote{\emph{Id}. For instance, if Alice
  possesses a plan to travel to a party at 9 p.m., then she can filter
  out all options that are incompatible with the possibility of
  attending the party, such as planning a dinner for 9 p.m. Faced with a
  smaller set of alternatives, Alice can more easily focus on the
  relevant questions: how to get to the party, and so on. And when 9
  p.m. arrives, Alice will, in fact, travel to the party, absent special
  circumstances.}

Because plans by their very nature strive to reduce the costs of
accomplishing one's ends, impossible, or inefficient, means are filtered
out. This is the feature of individual agency that allows law to dispose
individuals toward compliance. Those subject to law tend to believe that
(a) purposive agents, namely, legal officials, (b) create legal norms
addressed to them (c) in order to settle questions of how to
act.\footnote{Shapiro, \emph{supra} note 12, at 200.} If an individual
has these three beliefs, legal norms will occupy a privileged place in
an individual's deliberations.

Because such agents accept that a hierarchical relationship exists
between themselves and the state, such agents take the content of the
law to provide plans for their own actions and seek to ensure that other
plans are consistent with legal norms.\footnote{\emph{Id}., at 141.} The
content of legal norms thus has the capacity to induce rational agents
to integrate their conduct within the parameters of the law because such
agents \emph{take} the law to be an authority within its domain. Law
guides when its norms, owing to their status as dispositive settlers of
normative controversies, exert pressure on individuals to use their
content as the bases of plans. Law is designed to settle normative
debates. If law does what it claims to do, it will provide a stable
foundation for derivative plans. If the goals of a deliberative system
are to efficiently coordinate a number of desires, legal norms should
outperform any other type of norm, since no other type of norm guides
with the backing of the state. Individuals thus desire to incorporate
its content into their plans. This process of incorporation can occur
via either of two routes: filtering and facilitating.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Filtering and Facilitating}
\end{enumerate}

Legal norms that prohibit certain behavior filter out possible means for
the accomplishment of desires. For instance, if an individual has a
desire to ``get rich quick,'' criminal prohibitions on insider trading
might filter out the means of, inter alia, buying or selling securities
on material, nonpublic information in breach of a fiduciary
duty.\footnote{Such prohibitions may also filter out certain
  \emph{ends}. One might be enticed by the thought of a life of crime,
  for instance. Our focus on means here should not be read as denying
  that prohibitions also filter ends.} By filtering out certain
possibilities, non-filtered possibilities face fewer competitor means.
Legally sanctioned means are, in a sense, elevated as non-legal means
are winnowed away.

Yet, as Hart famously observed, legal norms do more than prohibit
conduct: they also create and facilitate new forms of conduct. By making
available new forms of action, legal norms create new means by which
individuals can accomplish their ends. Although it is odd to speak of
``compliance'' with legal norms that facilitate conduct, it is not odd
to imagine that legal officials in some sense favor the use of such ends
over the use of non-legal ends. It makes sense, then, to conceptualize
guidance as disposing individuals to employ legally-created forms of
action, even if the language of compliance is inapposite.

The content of legal norms filters out some means, and creates others.
The result is a set of legally sanctioned means. If individuals create
higher order intentions structured around this legally sanctioned set of
means, individuals will be more disposed to comply with the law. Law
influences dispositions by influencing the structural framework of --
the higher level plans -- individual deliberations. This dynamic has the
effect of crowding out alternative sources of planning norms. Recall
that legal norms compete against other norms to guide individual
conduct. One of the primary competitive advantages of law is that it
usually has the state's force behind it. Nonetheless, as the fact of
non-compliance shows, legal norms do not always claim victory. IG shows
how legal norms stay competitive by reducing the foothold that
alternatively-sourced norms might have in individuals' planning
mechanisms.

Raz hints at such an idea in his discussion of stability. ``{[}O{]}nly
if the law is stable are people guided by \emph{their knowledge of the
content of the law},'' Raz argues.\footnote{Raz, \emph{supra} note 13,
  at 229 (emphasis in original).} Implicit in Raz's argument is the idea
that, in the absence of legal guidance, individuals will plan according
to alternatively-sourced normative standards. If law is unstable,
individuals are guided by their knowledge of \emph{some other} norm.
Legal norms, though, can crowd out competing norms, and thus make them
less probable to influence individual behavior. By creating a set of
legally sanctioned means, law disposes individuals toward creating
higher order intentions that are compliant with law. The more
individuals craft their intentions in a legally compliant manner, the
more dependent they become upon legally compliant behaviors. As with
many systems engineered to produce certain outputs, the more engaged one
is with a system, the more dependent one becomes on that system, the
harder it becomes to exit that system. And so it is with law.

Plans allow individuals to coordinate vast swaths of desires in an
efficient manner. They are efficient solutions in part because they
allow individuals to incur the costs of weighing options, deciding on
means, and coordinating those choices with other of their plans and
other individuals. It thus seems as though plans only efficiently
structure decision-making if they are relatively stable. After all, ``if
we were constantly to be reconsidering the merits of our prior plans
they would be of little use in coordination and in helping us cope with
our resource limitations.''\footnote{Bratman, \emph{supra} note 56, at
  3.} ``Nonreconsideration,'' in other words, ``will typically be the
default,'' even in cases where, were the individual to reconsider the
decision, she would plan differently.\footnote{\emph{Id}.}

Under IG, legal norms must be capable of forming the basis of plans
through legally-sanctioned means. And plans, we just saw, are in the
normal course of things stable. Legal norms, then, too, it would seem,
must be stable. Predicating plans on unstable means seems like a recipe
for incurring the costs of reconsideration. The role of stability under
IG is, at bottom, about the way in which reconsideration affects
individuals' ability to construct plans around legally-sanctioned means.
Plans are premised on the existence of certain conditions. When the
conditions upon which a plan is premised change, an agent may reconsider
her previously formed plans, but she need not. Non-reconsideration is
the default position, but we can sketch with slightly more specificity
the conditions under which individuals override this default.

Changes in the world exact one of two types of costs in relation to
individual plans. On the one hand, changes can exact costs related to
the agent's deliberative structure. Reconsideration of plans premised on
changes takes time, energy, and inhibits coordination with others. If an
agent prefers to avoid those costs, he might incur a different type of
cost: the costs of attempting to bend a changed world back to a state in
which it is compatible with his original set of plans. For all
individuals interested to keep their plans and the state of the world in
some sense connected, change exacts one or the other of these types of
costs. Individuals choose different strategies to manage these
costs.\footnote{Bratman, \emph{supra} note 54.} At one end of the
spectrum, individuals' plans are only as strong as the set of beliefs
and desires that formed the original basis for the plan. If the agent's
plans are non-robust in the face change, then the agent will frequently
be attempting to bring his plans in line with the changed state of the
world. The costs typically borne of such an agent include lost
efficiency gains, since he expends resources recalculating means and
ends, and lost coordination gains, since others that rely on his
previous plans will no longer reliably do so.\footnote{For instance,
  consider the deliberative difficulties facing an agent with a higher
  order plan grounded in act consequentialism, or some other set of
  values that requires that an agent constantly calculate the value of
  her action.} Such plans will struggle to serve the usual purposes of a
plan, since they filter few actions out and ask the agent to continually
deliberate over new options. Nonetheless, some agents may hold such a
view, at least about some matters.

At the other end of the spectrum, plans possess an ``intrinsic
stability,'' which commits the agent to attempt to bring about their
satisfaction conditions, come what may.\footnote{\emph{Id}. at 11.} If
an agent never reconsiders his plans, then volatility will impact the
agent in a characteristic way; namely, the agent will be forced to
attempt to bring the world in line with her pre-established plans. The
costs typically borne of this type of agent include those involved with
the open-ended challenges of attempting to make whatever aspect of the
world is no longer compatible with his original set of plans once again
compatible. Outside of these two poles, individuals tolerate varying
levels of reconsideration. This variance need not be considered the
product of irrationality. Some individuals may be quite adept at
shouldering the costs of reconsideration, while others may be skilled at
bending the world to their wills.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Deliberative Types}
\end{enumerate}

We see, then, that individuals prefer to allocate change costs in
different ways. This allows us to imagine a spectrum of approaches to
the allocation of change costs.\footnote{This section draws heavily on
  Bratman's discussion of stability of intentions. \emph{See} Bratman,
  \emph{supra} note 56. Any given individual might be a certain type
  with regard to one issue, but a different type with regard to another
  issue, since the costs of being one type or the other will vary based
  on the issue and individual capacities.} We will identify two
intermediate types, producing four types in total.\footnote{We will
  assume that individuals know their deliberative types, and find it
  rational to be that type. If an individual knows his deliberative
  type, and the costs associated therewith, he will tend to structure
  his plans in such a way as to bring about the satisfaction of rational
  desires.} We can label these types as follows:

\begin{quote}
\textbf{1. Non-Robust}: Prior plans are always subject to
reconsideration, given a relevant change.

\textbf{2. Weakly Robust}: Prior plans are subject to reconsideration,
given a relevant change, only if the anticipated costs of
reconsideration are less than the anticipated benefits of
reconsideration, with regard to a specific action.

\textbf{3. Strongly Robust}: Prior plans are subject to reconsideration,
given a relevant change, only if the anticipated costs of
reconsideration are less than the anticipated benefits of
reconsideration, with regard to an agent's overall dispositions about
reconsideration.

\textbf{4. Intrinsically Stable}: Prior plans are never subject to
reconsideration, no matter the change.
\end{quote}

The existence of deliberative types complicates IG's account of the
value of planning. Recall that future-direct intentions, or plans, allow
us to deliberate more efficiently and coordinate intra- and
inter-personally.\footnote{Bratman, \emph{supra} note 55.} It would seem
to follow that an agent will not adopt or will tend not to adopt a plan
unless the plan facilitates efficient deliberation and intra- and
inter-personal coordination. Any plan, the content of which is likely to
induce inefficient deliberation or inhibit coordination is likely to
remain in an agent's ``plan library'' of possible, but unadopted,
plans.\footnote{\emph{Id}.} Given an agent's knowledge of her type, she
will rule out some plans in the plan library on grounds that they are
subject to too much volatility. Given a potential plan from the plan
library, there is some threshold of probability at which the agent,
given her tolerance for reconsideration, will reject the plan. Given the
choice of plan \emph{A}, there is some probability, Pr(\emph{A}), that,
out of the set of possible future events, at least one event that forces
the agent to reconsider will materialize. When that Pr(\emph{A}) reaches
the threshold that the agent sets based on her deliberative type, that
potential plan is rejected.

Strongly Robust and Intrinsically Stable types may be wary of
incorporating plans that they anticipate will be subject to volatility
that destabilizes those plans. This wariness may result from the fact
that they tend not to reconsider plans. Bending the world to one's plans
can be difficult, even where one prefers that course of action to
reconsidering her plans. Of course, some changes in the world may induce
such high costs that reconsideration is likely even for Strongly Robust
and Intrinsically Stable types. As a percentage of all changes, however,
these cases are likely to be low, given these types' policies about
change.

Strongly Robust and Intrinsically Stable types, precisely because they
make it a policy not to reconsider, may plan for contingencies. If this
is so, then these types may, in fact, \emph{not} be considerably more
wary of potentially unstable plans. Yet, contingency planning is costly,
just as buying insurance is costly. Thus, it will be sought to be
avoided where possible. In general, then, we can hypothesize that
Strongly Robust and Intrinsically Stable types will possess a
comparatively low volatility threshold. In other words, for these
deliberative types, a comparatively low level of volatility will cause
them to leave plans in their planning libraries, rather than adopt them.
For instance, say an Intrinsically Stable type is considering whether to
buy a piece of land on which sits a large lake, for the purposes of
fishing. An assessment of the regulatory climate, however, suggests that
fishing from lakes such as this one might soon be prohibited. The
question is, given our individual's policy of not reconsidering plans,
our individual decides not to buy the property. It seem as though,
because of his policy to not reconsider plans, our individual will
decide not to buy the property. But, as an Intrinsically Stable type, he
might be comfortable incurring the risk and attempting to bend the
regulatory process to his preferences. For symmetric reasons, Non-Robust
and Weakly Robust types will possess a comparatively high volatility
threshold.

Law's capacity to guide is typically thought of as a capacity that
extends uniformly over all individuals. IG requires a differentiated
approach to guidance, since it is premised on the idea that individuals
possess different tolerances for the risks associated with volatility.
By introducing Deliberative Types, we have introduced differences in the
population. We must find some way of discussing guidance as a capacity
possessed by law while also taking account of these individual
differences. The solution requires analysis of the dynamic and
cooperative character of legal guidance.

The state faces a limit on how often it can change the content of these
legal norms, if it desires individuals to use them as part of their
plans. Given a potential plan from the plan library, there is some
threshold of probability at which the agent, given her tolerance for
reconsideration, will reject the plan. This probability threshold,
however, is not permanently fixed or wholly independent of individuals'
wills, as we have thus far imagined.\footnote{Whether an individual is
  of a certain deliberative type is, in some sense, a function of that
  individual's desires. If an agent really is committed to act
  consequentialism, then it makes sense to treat prior plans as always
  or almost always subject to reconsideration. But whether an agent is
  in fact committed to act consequentialism is itself subject to
  reconsideration. Likewise, individuals can seek to reject fewer legal
  norms on grounds that they are insufficiently stable. An individual
  can lower the probability threshold at which he rejects possible legal
  norms due to its likely volatility if he moves closer to the
  Non-Robust type.} Even if we hold one's deliberative type constant --
even if one maintains that an individual's deliberative type is beyond
that individual's capacity to change -- the individual can, nonetheless,
structure his affairs such that the possibility of legal change is less
likely to reach the threshold beyond which the legal norm is perceived
to be too volatile to form the basis of a plan. In other words,
individuals can plan for legal change itself. But why do so? Purchasing
insurance, after all, is costly. Strongly Robust and Intrinsically
Stable types, precisely because they make it a policy not to reconsider
previously established plans, may choose to plan for certain sorts of
contingencies. If this is so, then these types may \emph{not} be
considerably more wary of potentially unstable plans. However, all
things equal, such planning is costly. Why incur these costs?

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Legal Guidance as Assurance Game}
\end{enumerate}

We can explain why individuals incur these costs by observing that the
state's provision of legal norms takes the form of an assurance game.
Both the state and individuals benefit when the state responds to widely
held problems \emph{and} individuals are disposed to adopt the
regulations. But both of these endeavors are costly, if unreciprocated.
One might hypothesize, then, that legal change will tend to occur when
both sets of parties are sufficiently assured that the other will
cooperate.

Let us assume that the state faces a choice between enacting a legal
norm that is responsive to some widespread problem. The state can either
change the content of the law or not change it. Citizens, meanwhile, can
choose to plan for volatility or not plan for it. Of course, some
individuals, namely, those Non-Robust types, need not plan for
volatility in order to cooperate. They are cooperative by their very
type. Outside of the special case of Non-Robust types, however, the
actions of the state and individuals are interdependent. The state is
better off changing legal norms in response to widespread problem if and
only if individuals tend to plan for volatility. Likewise, individuals
are better off incurring the costs of planning for volatility if and
only if the state enacts responsive legal regulations. Otherwise, they
have spent resources -- deliberative and perhaps material -- preparing
for legal guidance that never arrives.

We can represent legal guidance as an assurance game:

\begin{quote}
\emph{Individuals}

Change Not-Change
\end{quote}

\begin{longtable}[c]{@{}ll@{}}
\toprule
\emph{5}, \emph{5} & \emph{-5}, \emph{0}\tabularnewline
\midrule
\endhead
\emph{0}, \emph{-5} & \emph{0}, \emph{0}\tabularnewline
\bottomrule
\end{longtable}

\emph{State} Change

Not-Change

In order that the state be willing to change a legal norm, it likely
anticipates that individuals will be guided by it. As we observed
earlier, there is some volatility threshold above which individuals of a
certain deliberative type will not incorporate a legal norm into their
planning apparatus. Individuals encourage the state to enact new
regulations when they \emph{raise} that threshold.

An unreciprocated attempt at cooperation is worse than maintaining the
status quo. This is true of both the state and its subjects. If
individuals believe that the state cannot or will not enact responsive
regulation, it is preferable not to incur the costs of planning, all
other things equal. Likewise, the state will seek to avoid attempting to
provide legal guidance if it is likely to fail.

The dynamic conception of stability incorporates the possibility that
individuals will \emph{raise} their volatility thresholds. Although, as
we observed, raising one's volatility threshold is costly, we have now
seen why individuals may be inclined to do so. Individuals cooperate
with the state by planning for beneficial legal change. Where this is
true, volatility thresholds decrease. If so, then we have a solution to
our problem of differentiated guidance: volatility thresholds are no
longer so differentiated. If individuals plan for legal change, their
volatility thresholds clump together, thus providing officials with a
target range of stability that must be met. And, thus, we have derived
an answer to the question of why stability is a necessary condition in
order that law guide individual behavior. The degree of stability
necessary in order that law guide individuals is just that degree
required to meet the volatility threshold of some significant portion of
the population.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Conclusion}
\end{enumerate}

The conventional wisdom among positivists holds that instability
generates uncertainty and this uncertainty \emph{itself} impairs
guidance. In this paper, we have rejected that claim, and suggested,
instead, that individuals' \emph{expectations} about the volatility of
legal norms is far more important to legal guidance. If individuals
anticipate volatility, or have non-legal reasons to prefer volatility,
then unstable legal norms will pose few problems.

Raz argues that stability requires ``that the making of particular laws
should be guided by open and relatively stable general
rules.''\footnote{Raz, \emph{supra} note 13, at 213.} The concern is
that instability produces uncertainty, which results in individuals
failing to guide their conduct by law. When this is not the case, Raz
argues, it becomes ``difficult for people to plan ahead on the basis of
their knowledge of the law.''\footnote{\emph{Id}. at 216.} Our analysis
suggests a more complicated picture. Whether individuals are able to
plan ahead on the basis of probabilities about what the law will be
depends on their general deliberative type and the level of resources
that they have devoted to the issue. If the issue matters to
individuals, they will devote resources to understanding the
probabilities that the law will change; they will take out insurance to
compensate against change; and they will seek to push the law in one
direction or the other. What's more, as we have shown, individuals can
recognize that the state's responsiveness to issues can be a public
good, and respond to this realization by planning for legal change.

As one of the primary tools of democracy, law is deployed to respond to
problems. The sorts of problems that contemporary democratic regimes
solve do not arrive in neat intervals; they are often significant and
difficult, and the solutions partial and temporary. Instability is high
because long-term solutions are technically or politically impossible.
To demand stability in the name of legal guidance is to effectively
curtail the process of problem solving. Because it is usually preferable
that the state be flexible in its approach to problem solving, it is
important to know just how much instability is compatible with legal
guidance. We have tried to move past the simple refrain that law must be
stable, so common in the legal philosophy literature, and explain just
how much stability is, in fact, compatible with legal guidance.
